### 2023-10-08 feedback/20230927/article.tex
Organize feedback in folders

@@

There exists no standard definition of paradata (\cite{nicolaas2011survey,skold2022interrogating,huvila2022improving}).
Without declaring it a formal definition, however,
paradata can be understood in general sense as 'data about processes’ (with e.g. survey paradata
termed alternatively as records and audit trails) \cite{nicolaas2011survey},
or in more specific sense in relation to data collection, as 'data about the data collection process' (\cite{choumert2019using}).
This paper uses the latter as a working definition.

@@

There are multiple reasons why useful paradata matter. 
Most obvious is that having useful paradata gives an understanding how
data is produced.
This knowledge helps researchers from different fields to understand each other
and collaborate.
Additionally, useful open data is needed for Open Science to convincingly show
its benefits.
Finally, in computational fields, 
it can help understand how scholarly knowledge is produced (\cite{huvila2022improving}).
This paper discusses these general reasons applied to code.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate in which way code is paradata,
and why this type of paradata is relevant.
However, any field that uses computation 
and sensitive data in its experiments,
could be used as an example.

@@

Genetic epidemiology is a field within biology that
tries to determine the spread of heritable traits 
and their underlying biological mechanism.
For example, we know that lactose intolerance in adults is
caused by a decline in the production of lactose-degrading enzymes,
and is most commonly found in south-east Asia and south 
Africa (\cite{storhaug2017country}).
The trait is caused by the genetic make-up, or 'genotype'
\iffalse
(see Table \ref{tab:definitions} in the Supplementary Materials
for the definitions of terms used in this paper)
\fi
, of a person.
The trait, also called 'phenotype', 
in this example is lactose intolerance at adult age,
yet any human property, such as weight or height can be studied.
When an association between genotype and phenotype is found,
these associations, when relevant enough, are used to 
create so-called 'gene panels', 
where the location of the gene causing 
an association is measured specifically, 
to detect people at risk for the associated phenotype.
The rest of this section describes a genetic epidemiology study 
in more detail, with special focus on the computational experiment.

@@

This paper started with some suggestions to 
make paradata useful for data re(use):
paradata should be extensive, comprehensively documented,
with the creation of documentation and code going hand-in-hand,
as well as friendly to computers (\cite{huvila2022improving}).

@@

Dr. Bilderbeek has worked in computational biology for more than a decade.
Currently, he works for the National Bioinformatics Infrastructure Sweden (NBIS)
in Uppsala, Sweden, 
as an application expert and trainer, with a focus on reproducible research.
All his work can be found on GitHub at \url{https://github.com/richelbilderbeek}
and his YouTube channel.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{This paper}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This book chapter has been produced following all recommendations in it:
it's \LaTeX code is hosted on GitHub 
at \url{https://github.com/richelbilderbeek/chapter_paradata}.
Due this, one can, among others, see the complete version history of this paper.
This alone could allow research to be done on, for example,
the effect of reviewers' feedback on a first manuscript.
The website also supplies an alternative version of the same chapter
with more figures.

@@

\begin{table}[h]
  \begin{tabular}{|l|p{5cm}|p{5cm}|}
    \hline
    Term         & Definition                                                                   & Example                                                         \\
    \hline
    Code         & Body of text to be run by a computer                                         & The scripts run by a computational experiment                   \\
    Data         & Individual facts, statistics, or items of information                        & The name of a SNP known to have a significant association       \\
    Genotype     & The DNA allele at a certain location                                         & AA, AC, CG, GT, ...                                             \\
    Paradata     & Data that describe the process of generating data                            & The code to conclude that a SNP has a significant association   \\
    Phenotype    & How an organism looks like in the broadest sense                             & The concentration of IL6RA in the blood                         \\
    Primary data & Data that is the foundation of an experiment                                 & The genotype of individuals                                     \\
    Metadata     & Data that provide information about other data                               & The article that describes an experiment                        \\
    Raw data     & Data directly from the real world, that are not yet ready to be primary data & The raw results of getting the genotypes of individuals         \\
    SNP          & A named location within a genome                                             & \verb|rs12133641|                                               \\
    Trait        & A phenotype                                                                  & The concentration of IL6RA in the blood                         \\
    \hline
  \end{tabular}
  \caption{Terms used in this paper, and their definitions}
  \label{tab:definitions}
\end{table}

@@

\begin{table}[h]
  \begin{tabular}{|p{2cm}|l|}
    \hline
    \textbf{Type of code} & \textbf{Recommendation} \\
    \hline
    No code       & Disallow papers that do not supply code publicly \\
    \hline
    Raw code      & Don't publish raw code as text, host it on a code hosting website \\
    \hline
    Hosted code   & Use a standardised badge for build status \\
                  & Use a standardised badge for code coverage \\
                  & Standardise to link between paper and code host \\
    \hline
    Runnable code & Allow to upload containers without limits \\
                  & Website to host containers must allow for annotation \\
                  & Use a standardised badge for the container to be functioning  \\
                  & Standardise to link between container, paper and code repository \\
    \hline
  \end{tabular}
  \caption{Recommendations described in this paper}
  \label{tab:recommendations}
\end{table}

>>>>

% IH: replaced references to "this paper" with "this chapter". Check that ok.

@@

There exists no standard definition of paradata (\cite{nicolaas2011survey,skold2022interrogating,huvila2022improving}).
Without declaring it a formal definition, however,
paradata can be understood in general sense as 'data about processes’ (with e.g. survey paradata
termed alternatively as records and audit trails) \cite{nicolaas2011survey},
or in more specific sense in relation to data collection, as 'data about the data collection process' (\cite{choumert2019using}).
This chapter uses the latter as a working definition.

@@

There are multiple reasons why useful paradata matter. 
Most obvious is that having useful paradata gives an understanding how
data is produced.
This knowledge helps researchers from different fields to understand each other
and collaborate.
Additionally, useful open data is needed for Open Science to convincingly show
its benefits.
Finally, in computational fields, 
it can help understand how scholarly knowledge is produced (\cite{huvila2022improving}).
% IH: As the series editors asked that all chapters should in the end somehow link to knowledge management, may be here could be a place to add a few words about it e.g. see below
% IH: and to integrate this understanding as a part of the management of scholarly knowledge.
This chapter discusses these general reasons applied to code.

@@

This chapter uses genetic epidemiology as a specific example,
to illustrate in which way code is paradata,
and why this type of paradata is relevant.
However, any field that uses computation 
and sensitive data in its experiments,
could be used as an example.

@@

Genetic epidemiology is a field within biology that
tries to determine the spread of heritable traits 
and their underlying biological mechanism.
For example, we know that lactose intolerance in adults is
caused by a decline in the production of lactose-degrading enzymes,
and is most commonly found in south-east Asia and south 
Africa (\cite{storhaug2017country}).
The trait is caused by the genetic make-up, or 'genotype'
\iffalse
(see Table \ref{tab:definitions} in the Supplementary Materials
for the definitions of terms used in this chapter)
\fi
, of a person.
The trait, also called 'phenotype', 
in this example is lactose intolerance at adult age,
yet any human property, such as weight or height can be studied.
When an association between genotype and phenotype is found,
these associations, when relevant enough, are used to 
create so-called 'gene panels', 
where the location of the gene causing 
an association is measured specifically, 
to detect people at risk for the associated phenotype.
The rest of this section describes a genetic epidemiology study 
in more detail, with special focus on the computational experiment.

@@

This chapter started with some suggestions to 
make paradata useful for data re(use):
paradata should be extensive, comprehensively documented,
with the creation of documentation and code going hand-in-hand,
as well as friendly to computers (\cite{huvila2022improving}).

@@

% IH: Moved to a separate document
%Dr. Bilderbeek has worked in computational biology for more than a decade.
%Currently, he works for the National Bioinformatics Infrastructure Sweden (NBIS)
%in Uppsala, Sweden, 
%as an application expert and trainer, with a focus on reproducible research.
%All his work can be found on GitHub at \url{https://github.com/richelbilderbeek}
%and his YouTube channel.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{This chapter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This book chapter has been produced following all recommendations in it:
it's \LaTeX code is hosted on GitHub 
at \url{https://github.com/richelbilderbeek/chapter_paradata}.
Due this, one can, among others, see the complete version history of this chapter.
This alone could allow research to be done on, for example,
the effect of reviewers' feedback on a first manuscript.
The website also supplies an alternative version of the same chapter
with more figures.

@@

\begin{table}[h]
  \begin{tabular}{|l|p{5cm}|p{5cm}|}
    \hline
    Term         & Definition                                                                   & Example                                                         \\
    \hline
    Code         & Body of text to be run by a computer                                         & The scripts run by a computational experiment                   \\
    Data         & Individual facts, statistics, or items of information                        & The name of a SNP known to have a significant association       \\
    Genotype     & The DNA allele at a certain location                                         & AA, AC, CG, GT, ...                                             \\
    Paradata     & Data that describe the process of generating data                            & The code to conclude that a SNP has a significant association   \\
    Phenotype    & How an organism looks like in the broadest sense                             & The concentration of IL6RA in the blood                         \\
    Primary data & Data that is the foundation of an experiment                                 & The genotype of individuals                                     \\
    Metadata     & Data that provide information about other data                               & The article that describes an experiment                        \\
    Raw data     & Data directly from the real world, that are not yet ready to be primary data & The raw results of getting the genotypes of individuals         \\
    SNP          & A named location within a genome                                             & \verb|rs12133641|                                               \\
    Trait        & A phenotype                                                                  & The concentration of IL6RA in the blood                         \\
    \hline
  \end{tabular}
  \caption{Terms used in this chapter, and their definitions}
  \label{tab:definitions}
\end{table}

@@

\begin{table}[h]
  \begin{tabular}{|p{2cm}|l|}
    \hline
    \textbf{Type of code} & \textbf{Recommendation} \\
    \hline
    No code       & Disallow papers that do not supply code publicly \\
    \hline
    Raw code      & Don't publish raw code as text, host it on a code hosting website \\
    \hline
    Hosted code   & Use a standardised badge for build status \\
                  & Use a standardised badge for code coverage \\
                  & Standardise to link between paper and code host \\
    \hline
    Runnable code & Allow to upload containers without limits \\
                  & Website to host containers must allow for annotation \\
                  & Use a standardised badge for the container to be functioning  \\
                  & Standardise to link between container, paper and code repository \\
    \hline
  \end{tabular}
  \caption{Recommendations described in this chapter}
  \label{tab:recommendations}
\end{table}
